OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2018-01-08 manuscript_nar.tex
Answering R2 reviewer's comments

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content. This approach is used in the method of
\citet{yaffe2011probabilistic} and in HiCNorm by \citet{hu2012hicnorm}.
The second strategy is to implicitly correct unknown biases by enforcing
some regularity condition on the data. This approach is used in the
Iterative Correction and Eigenvector decomposition method (\textit{ICE})
of \citet{imakaev2012iterative}, whereby the total amount of contacts of
every bin is imposed to be the same. \textit{ICE} is currently the most
popular method, due in part to its speed.

@@

So far, the only attempt to address the issue was the chromosome-adjusted
Iterative Correction Bias method (\textit{caICB}) of
\citet{wu2016computational}. However, \textit{caICB} applies a uniform
chromosome-wide copy number correction, effectively excluding the numerous
cases of partial aneuploidy and regional copy number variations.

@@

The most common representation of Hi-C data is a contact matrix, obtained
by slicing the genome in $n$ consecutive bins of fixed size (the
resolution) and computing the number of contacts between each pair of
bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
quantifying the interaction between the two loci at positions $i$ and
$j$.

@@

The Hi-C experiments were performed in T47D breast cancer cell lines
(aberrant karyotype, six samples), K562 leukemia cell lines (aberrant
karyotype, eight samples), mouse primary B cells (normal karyotyp, six
samples) ES cells (normal karotype, seven samples), GM12878
B-lymphoblastoid cells (normal karyotype, fifty eight samples), BT474
breast cancer cell line (aberrant karyotype, one sample), MCF10 breast
cancer cell line (aberrant karyotype, one sample), MCF-7 cancer cell line
(aberrant karyotype, one sample) and SKBR3 breast cell lines (aberrant
karyotype, one sample).

@@

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 or mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks} (for more details see BioRxiv:
https://doi.org/10.1101/132456). In addition, the pipeline is available
from the Supplementary Material published by \citet{quilez2017managing}.

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720, see Supplementary Table~1), that were performed
in either B cells or ES cells, and from either CRG or UCSC. The simulation
strategy was the following: First we selected uniformly at random the
amount of copy number break points (from 3 to 10). We then placed the
break points along chromosomes 18 and 19 uniformly at random. Next, we
assigned a copy number (2, 3, 4 or 10 with equal probability) to each
segment delimited by the break points. We computed the outer product of
this simulated copy number profile and multiplied it element-wise with the
original contact matrices of chromosomes 18 and 19. The resulting matrices
were used as input for correction methods. For each simulation (100 in
total), we measured the pairwise reproducibility score defined by
\citet{yan2017hicspector} and computed the average for pairs from the same
cell type minus the average for pairs from different cell types.

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mbp in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\citet{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\citet{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The purple line represents the raw signal, the dark blue
line represents the signal after OneD bias correction, the light blue line
represents the signal after OneD bias and copy number (CN) correction and
the orange line represents the signal after ICE bias correction. The long arm
of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
four copies, explaining that the signal is about twice higher than for
the short arm.}
\label{fig:totals}
\end{figure}

@@

We used four metrics to compare Hi-C matrices (see Materials and Methods).
For consistency and concision, the results based on the reproducibility
score of \citet{yan2017hicspector} are shown in the main figures, and the
results based on the other metrics are shown in the Supplementary
Material.

@@

To test this idea, we collected a Hi-C data set from breast cancer cell
lines and measured the reproducibility score between identical cell types
after normalization with the different tools. In this case, we found that
\textit{OneD+CN} outperformed \textit{OneD} and gave the highest overall
reproducibility (Figure~\ref{fig:cnv_correction}A). Even though the
variability is high in this case, this suggests that the karyotypes of
these cells have diverged, so that the copy number correction performed by
\textit{OneD} increased the reproducibility between samples from the same
cell type.

@@

Copy number correction is also useful to give euploid-equivalent
representations of samples from aberrant karyotypes.
Figure~\ref{fig:cnv_correction}B shows an example where a region at the
center of the matrix is duplicated. \textit{ICE} overcompensated the
original bias and faded the signal almost entirely. Concomitantly, the
signal at the bottom left of the matrix was enhanced and showed a
structure that was not visible in the raw data. On the contrary,
\textit{LGF} strengthened the central region and the diagonal.
\textit{OneD} reduced the level of the central portion by a factor 2
approximately, but did not otherwise distort the main features of the
region.

@@

We reasoned that \textit{OneD} should handle such cases gracefully,
because the signal is formally equivalent to an aberrant karyotype where
most of the genome is present in zero copy. We thus evaluated the
performance of \textit{OneD} on a Capture Hi-C data set at 5~Kbp
resolution centered on a 6~Mbp domain of chromosome 6 in T47D cells. We
found that the reproducibility score between replicates was higher after
normalization with \textit{OneD} than with matrix-balancing methods
(figure~\ref{fig:chic}A). The performance of \textit{OneD} was only
slightly lower than the raw data, suggesting that it barely disturbs the
biological signal. Other tools were not included beause they had
prohibitive run time.

@@

\textit{OneD} corrects a single variable instead of the whole matrix, and
thus estimates the model parameters much faster than previous explicit
methods. We measured the running time of the different tools on a 3.3~GHz
machine with 62 GB~RAM, always using the default parameters.
Figure~\ref{fig:times}A shows the running times of the different methods
on all the samples shown in Supplementary Figure~1 at 100~Kbp resolution.
The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
an over 100-fold span between the two.

@@

To better understand these results, we also investigated the influence of
bin size and sequencing depth on the running time. We performed a
benchmark on the sample with highest sequencing depth in the data set
(HIC070, see Supplementary Table~1) at different bin sizes while
downsampling the contact matrices to mimic lower resolutions. On this
test, we included only the methods competing with \textit{OneD} in terms
of speed, \textit{i.e.} the matrix-balancing methods.
Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
nearly identical results at very different sequencing depths
(Supplementary Figure~13). \textit{vanilla} was the overall fastest method
and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
at high but not at low sequencing depth. At low resolution (100~Kbp) the
speed advantage of \textit{OneD} appeared from low sequencing depth, but
at high resolution resolution (5~Kbp) it appeared only at high sequencing
depth.

>>>>

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content. This approach is used in the method of
\cite{yaffe2011probabilistic} and in HiCNorm by \cite{hu2012hicnorm}.
The second strategy is to implicitly correct unknown biases by enforcing
some regularity condition on the data. This approach is used in the
Iterative Correction and Eigenvector decomposition method (\textit{ICE})
of \cite{imakaev2012iterative}, whereby the total amount of contacts of
every bin is imposed to be the same. \textit{ICE} is currently the most
popular method, due in part to its speed and simplicity.

@@

So far, the only attempt to address the issue was the chromosome-adjusted
Iterative Correction Bias method (\textit{caICB}) of
\cite{wu2016computational}. However, \textit{caICB} applies a uniform
chromosome-wide copy number correction, effectively excluding the numerous
cases of partial aneuploidy and regional copy number variations.

@@

The most common representation of Hi-C data is a contact matrix, obtained
by slicing the genome in $n$ consecutive bins of fixed size (the
resolution) and computing the number of contacts between each pair of
bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
quantifying the interaction frequency between the two loci at positions
$i$ and $j$.

@@

The Hi-C experiments were performed in T47D breast cancer cell lines
(aberrant karyotype, six samples), K562 leukemia cell lines (aberrant
karyotype, eight samples), mouse primary B cells (normal karyotype, six
samples) ES cells (normal karotype, seven samples), GM12878
B-lymphoblastoid cells (normal karyotype, fifty eight samples), BT474
breast cancer cell line (aberrant karyotype, one sample), MCF10 breast
cancer cell line (aberrant karyotype, one sample), MCF-7 cancer cell line
(aberrant karyotype, one sample) and SKBR3 breast cell lines (aberrant
karyotype, one sample).

@@

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 or mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks} (for more details see BioRxiv:
https://doi.org/10.1101/132456). In addition, the pipeline is available
from the Supplementary Material published by \cite{quilez2017managing}.

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720, see Supplementary Table~1), that were performed
in either B cells or ES cells, and from either CRG or UCSD. The simulation
strategy was the following: First we selected uniformly at random the
amount of copy number break points (from 3 to 10). We then placed the
break points along chromosomes 18 and 19 uniformly at random. Next, we
assigned a copy number (2, 3, 4 or 10 with equal probability) to each
segment delimited by the break points. We computed the outer product of
this simulated copy number profile and multiplied it element-wise with the
original contact matrices of chromosomes 18 and 19. The resulting matrices
were used as input for correction methods. For each simulation (100 in
total), we measured the pairwise reproducibility score defined by
\cite{yan2017hicspector} and computed the average for pairs from the same
cell type minus the average for pairs from different cell types.

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mbp in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\cite{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The purple line represents the raw signal, the dark blue
line represents the signal after OneD bias correction, the light blue line
represents the signal after OneD bias and copy number (CN) correction and
the orange line represents the signal after ICE bias correction. The long arm
of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
four copies, explaining that the signal is about twice higher than for
the short arm. Both figures present data at 100~Kbp resolution}
\label{fig:totals}
\end{figure}

@@

We used four metrics to compare Hi-C matrices (see Materials and Methods).
For consistency and concision, the results based on the reproducibility
score of \cite{yan2017hicspector} are shown in the main figures, and the
results based on the other metrics are shown in the Supplementary
Material.

@@

To test this idea, we collected a Hi-C data set from breast cancer cell
lines and measured the reproducibility score between identical cell types
after normalization with the different tools. In this case, we found that
\textit{OneD+CN} outperformed \textit{OneD} and gave the highest overall
reproducibility (Figure~\ref{fig:cnv_correction}A). Even though the
variability is high in this case, this suggests that the karyotypes of
these cells have diverged, so that the copy number correction performed by
\textit{OneD+CN} increased the reproducibility between samples from the same
cell type.

@@

Copy number correction is also useful to give euploid-equivalent
representations of samples from aberrant karyotypes.
Figure~\ref{fig:cnv_correction}B shows an example where a region at the
center of the matrix is duplicated. \textit{ICE} overcompensated the
original bias and faded the signal almost entirely. Concomitantly, the
signal at the bottom left of the matrix was enhanced and showed a
structure that was not visible in the raw data. On the contrary,
\textit{LGF} strengthened the central region and the diagonal.
\textit{OneD+CN} reduced the level of the central portion by a factor 2
approximately, but did not otherwise distort the main features of the
region.

@@

We reasoned that \textit{OneD} should handle such cases gracefully,
because the signal is formally equivalent to an aberrant karyotype where
most of the genome is present in zero copy. We thus evaluated the
performance of \textit{OneD} on a Capture Hi-C data set at 5~Kbp
resolution centered on a 6~Mbp domain of chromosome 6 in T47D cells. We
found that the reproducibility score between replicates was higher after
normalization with \textit{OneD} than with matrix-balancing methods
(figure~\ref{fig:chic}A and Supplementary Figure~13). The performance
of \textit{OneD} was only slightly lower than the raw data, suggesting
that it barely disturbs the biological signal. Other tools were not
included beause they had prohibitive run time.

@@

\textit{OneD} corrects a single variable instead of the whole matrix, and
thus estimates the model parameters much faster than previous explicit
methods. We measured the running time of the different tools on a 3.3~GHz
machine with 62 GB~RAM, always using the default parameters.
Figure~\ref{fig:times}A shows the running times of the different methods
on all the samples shown in Supplementary Table~1 at 100~Kbp resolution.
The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
an over 100-fold span between the two.

@@

To better understand these results, we also investigated the influence of
bin size and sequencing depth on the running time. We performed a
benchmark on the sample with highest sequencing depth in the data set
(HIC070, see Supplementary Table~1) at different bin sizes while
downsampling the contact matrices to mimic lower depths. On this
test, we included only the methods competing with \textit{OneD} in terms
of speed, \textit{i.e.} the matrix-balancing methods.
Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
nearly identical results at very different sequencing depths
(Supplementary Figure~14). \textit{vanilla} was the overall fastest method
and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
at high but not at low sequencing depth. At low resolution (100~Kbp) the
speed advantage of \textit{OneD} appeared from low sequencing depth, but
at high resolution resolution (5~Kbp) it appeared only at high sequencing
depth.

### 2018-01-08 supplementary_figures_nar.tex
Answering R2 reviewer's comments

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_13.eps}}
    \caption{Correlation between subsamples at different resolutions and
    depths. X axis: sequencing depth. Y axis: Pearson correlation between the
    correction vectors of two different subsamples. Each color represents a
    correction method and each panel a bining resolution. Note the Y axis limits
    are specific to each panel.}
\end{figure}

>>>>

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_13.eps}}
    \caption{Bias removal on Capture Hi-C data A and C. Performance on the Capture
    Hi-C data. Two replicates centered on a 6~Mbp domain on chromosome 6 in
    T47D cells were normalized by the indicated tools and their
    Pearson (A) or Spearman (C) correlation computed. B and D. Performance on virtual
    Capture Hi-C  data. Two Hi-C experiments were post-processed to produce an output
    similar to the Capture Hi-C experiment depicted in panels A and B. The signal
    outside the domain on chromosome 6 was removed and the data were
    normalized (capt), or the data were normalized before removing the signal
    outside the domain on chromosome 6 (full). Four matrices per method were
    generated and the Pearson (A) or Spearman (C) correlation computed between the
    pairs from different replicates were computed. Note that the scores are identical
    for the raw data because in this case the matrices of the same replicate are
    identical.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_14.eps}}
    \caption{Correlation between subsamples at different resolutions and
    depths. X axis: sequencing depth. Y axis: Pearson correlation between the
    correction vectors of two different subsamples. Each color represents a
    correction method and each panel a bining resolution. Note the Y axis limits
    are specific to each panel.}
\end{figure}
